%!TEX encoding = UTF-8 Unicode

%% Thesis cover pages
%%
%% Based on the example.tex file provided by Tomás Oliveira e Silva
%%
%% Modified by João Paulo Barraca <jpbarraca@ua.pt>
%% Modified by Filipe Manco <filipe.manco@ua.pt>

% Remove 'coverdepname' parameter to remove department name from cover pages
% Add 'final' parameter to remove "DOCUMENTO PROVISÓRIO" from cover page
\documentclass[11pt,twoside,a4paper,final]{report}

%%% This file is a template for the cover of your
%%%  thesis.
%%% Lines like this one, starting with %%% , are
%%%  instructions. Search them to know where and
%%%  how to edit this file. After editing the
%%%  document you can delete them.
%%% Items marked with OPTIONAL can be removed.
%%% Most of the lines you will have to change
%%%  already have examples to help you understand.
%%% The complete list of options and commands can
%%%  be found in uaThesisTemplate.sty


%%% Select your area of study and department from
%%%  the following lists and replace them in the
%%%  \usepackage declaration ([area,department]).
%%%
%%% | area        |   | department |
%%% ---------------   --------------
%%% | arts        |   | dao        |
%%% | sciences    |   | dbio       |
%%% | education   |   | dcspt      |
%%% | economy     |   | deca       |
%%% | engineering |   | degeit     |
%%% | humanities  |   | de         |
%%% | health      |   | deti       |
%%% |             |   | demac      |
%%% |             |   | decivil    |
%%% |             |   | dem        |
%%% |             |   | dfis        |
%%% |             |   | dlc        |
%%% |             |   | dmat       |
%%% |             |   | dq         |
%%% |             |   | dcm        |
%%% |             |   | dgeo        |
%%% ---------------   --------------
%% Import the UA thesis template
\usepackage[health,dcm]{uaThesisTemplate}


%% Optional packages for utf8 encoding and T1 fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

%% For prettier pages
\usepackage{lmodern}
\usepackage{hyperref}
%% We will use english and portuguese
\usepackage[english,portuguese]{babel}

\usepackage{xspace}% used by \sigla

%%% This package is only used to create a dummy
%%%  text for the abstract. You can remove the
%%%  following line in the final document.
%\usepackage{blindtext}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%       DEFINE THESIS COVER FIELDS
%%                  START
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Edit thesis year
%% Thesis Year
\ThesisYear{\the\year}


%%% Uncomment one of the following options:
\DocumentTypeThesis
%\DocumentTypeThesisProposal
%\DocumentTypeDissertation

%%% Uncomment if you are a student of a MAP Doctoral program 
%%% http://www.map.edu.pt/
%%% | area                | identifier |
%%% | Biology of Plants   | bioplant   |
%%% | Computer Science    | i          |
%%% | Physics             | fis        |
%%% | Telecommunications  | tele       |

%\MAPProgram{i}

%%% Edit author's name
%% Author's full name
\Author{Pedro Filipe\\ Carneiro Venâncio}

%%% Edit thesis title in Portuguese
%% Thesis Mandatory Title in Portuguese
\TitlePT{Aplicação de novos algoritmos bioinformáticos na análise de dados de Next Generation Sequencing (NGS)}

%%% OPTIONAL: you can write your thesis title in
%%%  other languages. You can repeat the following
%%%  line as many times as you want, or remove it
%%%  if you only want the portuguese title.
%% Optional Titles in additional languages.
\TitleEN{Application of new bioinformatic algorithms in Next Generation Sequencing (NGS) data analysis.}

%%% OPTIONAL: add a quotation if you like.
%% Optional Quotation
\Quote{\selectlanguage{english}Sometimes it's the people no one can imagine anything of who do the things no one can imagine}{Alan Turing, mathematician and computer scientist}

%%% Edit your supervisor's information.
%%%
%%% Use this command for male supervisor
%%% \Supervisor{name}{category}
%%%
%%% Or this command for femmale supervisor
%%% \Supervisora{name}{category}
%%%
%%% name   : Supervisor's name
%%% categoy: One of the following list
%%%
%%% | category                  |
%%% -----------------------------
%%% | catedrático/a             |
%%% | catedrático/a convidado/a |
%%% | associado/a c/ agregação  |
%%% | associado/a               |
%%% | auxiliar                  |
%%% | auxiliar convidado        |
%%% -----------------------------
%% Supervisor information
\Supervisora{Gabriela Maria Ferreira Ribeiro de Moura}{auxiliar}

%%% OPTIONAL: you can add a second supervisor.
%%%
%%% Use this command for male supervisor
%%% \SecondSupervisor{name}{category}
%%% 
%%% Or this command for femmale supervisor
%%% \SecondSupervisora{name}{category}
%%%  {[preposition: (at, in, no, da)] department}
%%%  {[preposition: (at, in, do, na)] school}
%%%
%%% name      : See above.
%%% category  : See above.
%%% department: Can be a department, faculty or
%%%             other; Choose the preposition
%%%             accordingly.
%%% school    : University/Scholl name; Choose the
%%%             preposition accordingly.
%% Optional SecondSuperviser information
\SecondSupervisora{Alexandra Filipa Lopes}{membro da entidade de acolhimento}{Unilabs}

%%% Edit your degree information.
%%%
%%% \Degree{title}{name}
%%%
%%% title: Degree title (Mestre or Doutor)
%%% name : Degree name
%% Degree title and name
\Degree{Mestre}{Bioinformática Clínica, especialização em Bioinformática do Genoma}

%%% OPTIONAL: edit grant support texts where
%%%  appropriate. Maximum two.
%%%
%%% \GrantText{grant1}{grant2}
%% Optional Grant support texts
%%\GrantText{Texto Apoio financeiro do POCTI no âmbito do III Quadro Comunitário de Apoio.}
%%          {Texto Apoio financeiro da FCT e do FSE no âmbito do III Quadro Comunitário de Apoio.}

%%% OPTIONAL: Add a dedication text.
%% Optional Dedication text
\Dedication{Dedico este trabalho à Bárbara e à minha família e amigos.}

%%% OPTIONAL: Add an acknowledgement text
%% Optional Acknowledgement text
\Acknowledgements{Agradeço à Alexandra Lopes, a minha orientadora na Unilabs, pela confiança que sempre demostrou no meu trabalho. A tua disponibilidade constante e a forma prática como me foste guiando ao longo deste percuso foram essenciais para chegar até aqui.

À Professora Gabriela Moura, a minha orientadora na Universidade de Aveiro, agradeço por me ter colocado no caminho certo e por ter acreditado em mim e neste projeto. A sua orientação foi fundamental para manter o foco e a direção certa.

Ao Professor Bruno Gago, Diretor do Mestrado, agradeço por nunca ter desistido de mim, mesmo quando o caminho parecia impossível. Obrigado por ter sido um guia e uma referência.

Ao Professor Miguel Pais Vieira, agradeço pela forma como me preparou ao longo deste ano para este desafio. A sua exigência e rigor foram essenciais para que este projeto fosse um sucesso. Obrigado por ter acreditado em mim.

A todos os professores com quem me cruzei. Obrigado por terem partilhado o vosso conhecimento e experiência, por terem sido uma inspiração e por terem contribuído para a minha formação.

Ao Ricardo Pais, um obrigado especial pela persistência e pela paciência. Obrigado por ter estado sempre disponível para ajudar e por ter sido um verdadeiro salvavidas a resolver cada problema que foi surgindo (e foram muitos).

À Béryl Royer-Bertrand, por, mesmo à distância, ter dado um contributo essencial através do seu vasto conhecimento em bioinformática. A sua ajuda foi crucial em momentos chave, e sou grato por isso.

Ao Alberto Pessoa, meu colega e amigo, um grande obrigado pelo apoio, pelas discussões francas e pelas críticas construtivas que ajudaram a dar forma a este trabalho.

Aos meus colegas da Unilabs, obrigado por me terem acolhido tão bem e por tornarem esta experiência mais leve e enriquecedora. O ambiente de partilha e camaradagem foi essencial para que este percurso fosse tão positivo.

Aos meus amigos de Bragança e de Aveiro, agradeço por estarem sempre presentes, por me apoiarem quando foi preciso e por nunca deixarem de acreditar nesta amizade, mesmo quando estive mais ausente.

À família que me adotou através da Bárbara, o meu obrigado pelo carinho e apoio incondicional. Vocês tornaram este caminho muito mais fácil e agradável com a vossa amizade e alegria.

À minha família, ao meu pai, à minha mãe e às minhas avós, pela oportunidade que me deram e por nunca desistirem de mim, mesmo quando parecia que o melhor era desistir. A eles devo a educação e a pessoa em que me tornei. Obrigado pelo sacrifício. Valeu a pena!

E, por fim, à Bárbara. Obrigado por estares sempre ao meu lado, por seres o meu apoio nos dias bons e nos dias menos bons. Deste-me força quando precisei, foste a voz da razão quando perdi o norte e o ombro onde me apoiei em todas as fases deste percurso. Sem ti, este projeto teria sido muito mais difícil. Esta conquista também é tua.
}

%%% Edit the keywords and abstract in Portuguese
%%% Don't forget to remove the \blindtext[3]
%%%  declaration.
%%% Don't remove the \selectlanguage statement or you will end with incorrect hyphenation
%% Mandatory Keywords and Abstract in Portuguese
\Abstract{Palavras Chave}{Sequenciação de Nova Geração (NGS), Bioinformática, Análise de Dados Genómicos, Profundidade de Cobertura, Amplitude de Cobertura, Samtools, Painéis de Genes, Exoma, Gene Único}
         {Resumo}{\selectlanguage{portuguese}Os avanços na Sequenciação de Nova Geração (NGS) transformaram a medicina genómica, permitindo uma análise de dados genéticos rápida e acessível. No entanto, a deteção precisa de variantes genéticas e a avaliação da qualidade da sequenciação continuam a ser desafios no diagnóstico clínico. Esta trabalho apresenta o desenvolvimento de um software que automatiza o cálculo de métricas chave de NGS, como a profundidade de cobertura e a percentagem de cobertura em diversos limiares. O software utiliza ferramentas bioinformáticas como SAMtools e integra uma interface gráfica desenvolvida em Streamlit, permitindo a análise de painéis de genes, exoma e genes individuais, com flexibilidade na configuração e exportação de resultados.\\Testado em grandes conjuntos de dados genómicos, o software calculou com sucesso as métricas necessárias, proporcionando informações detalhadas sobre a qualidade dos dados. A ferramenta demonstrou eficácia na identificação de regiões de baixa cobertura, facilitando a deteção de variantes genéticas e melhorando a acessibilidade para clínicos. Ao garantir uma cobertura abrangente das regiões-alvo, o software aumenta a fiabilidade na análise de NGS, apoiando diagnósticos mais precisos de doenças genéticas e raras.\\\\Link para o repositório do projeto: \url{https://github.com/pfcv99/metrics_app.git}}
%%% OPTIONAL: you can write the keywords and
%%%  abstract in other languages. You can repeat
%%%  the following lines as many times as you want
%%%  or remove them if you only want portuguese.
%%% Don't remove the \selectlanguage statement or you will end with incorrect hyphenation.
%%% You can modify the language to any other supported by Latex
%% Optional Keywords and Abstract in additional languages.
\Abstract{Keywords}{Next Generation Sequencing (NGS), Bioinformatics, Genomic Data Analysis, Depth of Coverage, Breadth of Coverage, Samtools, Gene Panels, Exome, Single Gene}
         {Abstract}{\selectlanguage{english}Advancements in Next Generation Sequencing (NGS) have transformed genomic medicine, enabling rapid and accessible analysis of genetic data. However, the accurate detection of genetic variants and the assessment of sequencing quality remain challenges in clinical diagnostics. This work presents the development of software that automates the calculation of key NGS metrics, such as depth of coverage and percentage of coverage at various thresholds. The software utilizes bioinformatics tools like SAMtools and integrates a graphical interface developed in Streamlit, allowing for the analysis of gene panels, exomes, and individual genes, with flexibility in configuration and result exportation.\\Tested on large genomic datasets, the software successfully calculated the required metrics, providing detailed insights into data quality. The tool demonstrated effectiveness in identifying low coverage regions, facilitating the detection of genetic variants and improving accessibility for clinicians. By ensuring comprehensive coverage of target regions, the software enhances reliability in NGS analysis, supporting more accurate diagnoses of genetic and rare diseases. \\\\Link to the project repository: \url{https://github.com/pfcv99/metrics_app.git}}

\AI{Acknowledgement of use of AI tools}{\textbf{Recognition of the use of generative Artificial Intelligence technologies and tools, software and other support tools.}}
         {}{\selectlanguage{english}I acknowledge the use of ChatGPT 3.5 (Open AI, https://chat.openai.com) for paraphrasing and translations, the use of Adobe Illustrator (Adobe, https://www.adobe.com/pt/products/illustrator) for image creation and editing, the use of FreePik (FreePik, https://br.freepik.com/) for vector images download, and the use of diagrams.net (diagrams.net, https://app.diagrams.net/) for diagram creation.
         }

%%% Edit the Jury President information.
%% First Member will be the President of the Jury
\JuriMember{Professor Doutor Miguel Santos Pais Vieira}{Professor Auxiliar em Regime Laboral, Universidade de Aveiro}

%%% Repeat the above declaration for additional
%%%  members of the jury as many times as necessary.
%% Additional members of the Jury
\JuriMember{Doutor Miguel Monsanto Pinheiro}{Programador Senior, Critical Techworks}
\JuriMember{Professora Doutora Gabriela Maria Ferreira Ribeiro de Moura}{Professora Associada, Universidade de Aveiro}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%       DEFINE THESIS COVER FIELDS
%%                  END
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\begin{document}

%%% By default we consider portuguese to be the
%%%  main language (the last one in the usepackage
%%%  declaration). If you're writing your thesis
%%%  in english uncomment the following line.
%%% Set the proper language or you will end with incorrect hyphenation!
\selectlanguage{english}


\MakeCover
\end{document}
